---
title: "Statistical Rethinking 2: Chapter 5"
author: Vincent Arel-Bundock
---

```{r, message=FALSE}
source('helper.R')
```

# Section 5.1

```{r}
standardize <- function(x) as.vector(scale(x))
stan_data <- read.csv('../rethinking/data/WaffleDivorce.csv', sep = ';') %>%
             mutate(across(c(Divorce, Marriage, MedianAgeMarriage), standardize)) %>%
             compose_data
```

#### Plausible regression lines implied by the priors

We will estimate a series of regression models with a constant $\alpha$ and regression coefficients $\beta_k$, and these priors:

$$\alpha \sim N(0, .2)$$
$$\beta_k \sim N(0, .5)$$

To see if these priors make sense, we can plot a few of the regression lines implied by these priors. To do this, we draw random numbers from the distributions above, and we plot the corresponding regression lines:

```{r}
a = rnorm(50, 0, .2)
b = rnorm(50, 0, .5)
p <- ggplot()
for (i in 1:50) {
    p <- p + geom_abline(slope = b[i], intercept = a[i])
}
p + xlim(-3, 3) + ylim(-3, 3) +
    labs(x = 'Median age marriage (std)',
         y = 'Divorce rate (std)')
```

# Linear regression model in Stan

```{r}
model <- '
data {
  int<lower=1> N;   // number of observations
  int<lower=1> K;   // number of regressors (including constant)
  vector[N] y;      // outcome
  matrix[N, K] X;   // regressors
}
parameters {
  real<lower=0,upper=50> sigma;    // scale
  vector[K] b;                     // coefficients (including constant)
}
transformed parameters {
  vector[N] mu;                    // location
  mu = X * b;
}
model {
  y ~ normal(mu, sigma);    // probability model
  sigma ~ exponential(1);   // prior for scale
  b[1] ~ normal(0, 0.2);    // prior for intercept
  for (i in 2:K) {          // priors for coefficients
    b[i] ~ normal(0, 0.5);
  }
}
generated quantities {
  vector[N] yhat;           // predicted outcome
  for (i in 1:N) yhat[i] = normal_rng(mu[i], sigma);
}
'

mod1 <- fit_model(stan_program, stan_data)
```

#### Regression lines implied by the posterior

```{r}
datplot <- extract(mod1, 'b')
idx <- sample(1:nrow(datplot$b), 1000)

p <- ggplot()
for (i in idx) {
    intercept <- datplot$b[i, 1]
    slope <- datplot$b[i, 2]
    p <- p + geom_abline(intercept = intercept, slope = slope, alpha = .02)
}
p + xlim(-3, 3) +
    ylim(-3, 3) +
    labs(x = 'Median age marriage (std)',
         y = 'Divorce rate (std)')
```

# Alternative models

```{r}
dat2 <- prepare_data(Divorce ~ Marriage, dat)
dat3 <- prepare_data(Divorce ~ Marriage + MedianAgeMarriage, dat)

mod2 <- stan(model_code = model, data = dat2)
mod3 <- stan(model_code = model, data = dat3)

print(mod1, c('b', 'sigma'))
print(mod2, c('b', 'sigma'))
print(mod3, c('b', 'sigma'))
```

# Posterior predictive check (Figure 5.5)

```{r, warning=FALSE}
posterior <- extract(mod3, 'mu')$mu
datplot <- dat %>%
           mutate(`Predicted divorce` = colMeans(posterior),
                  lb = apply(posterior, 2, quantile, probs = .025),
                  ub = apply(posterior, 2, quantile, probs = .975),
                  label = dat$Loc)

ggplot(datplot, aes(Divorce, `Predicted divorce`, ymin = lb, ymax = ub, label = label)) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed') +
  geom_linerange(alpha = .3) +
  geom_text() +
  xlim(-2.5, 2.5) + ylim(-2.5, 2.5)
```

# Section 5.3

## Height (R code 5.48)

```{r, message=FALSE}
dat <- read.csv('../rethinking/data/Howell1.csv', sep = ';') %>%
       mutate(sex = ifelse(male == 1, 2, 1)) %>%
       list('height' = .$height, 'sex' = .$sex, 'N' = nrow(.), weight = .$weight)

model <- '
data {
  int<lower=1> N;        // number of observations
  vector[N] height;      // outcome
  int sex[N];            // regressor
}
parameters {
  real<lower=0,upper=50> sigma;    // scale
  vector[2] a;
}
transformed parameters {
  vector[N] mu;                    // location
  mu = a[sex];
}
model {
  height ~ normal(mu, sigma);    // probability model
  sigma ~ uniform(0, 50);        // prior: scale
  a[1] ~ normal(178, 20);        // prior: women
  a[2] ~ normal(178, 20);        // prior: men
}
generated quantities{
  real diff_fm;
  diff_fm = a[1] - a[2];
}
'
mod <- stan(model_code = model, data = dat)

summary(mod, c('a', 'diff_fm'))$summary
```

# Milk (R code 5.52)

```{r, message=FALSE}
dat <- read.csv('../rethinking/data/milk.csv', sep = ';') %>%
       mutate(clade_id = as.integer(as.factor(clade)), K = as.vector(scale(kcal.per.g))) %>%
       select(K, clade_id) %>%
       as.list
dat$N <- length(dat$K)
      
model <- '
data {
  int<lower=1> N;        // number of observations
  vector[N] K;           // outcome
  int clade_id[N];
}
parameters {
  real<lower=0,upper=50> sigma;    // scale
  vector[4] a;
}
transformed parameters {
  vector[N] mu;                    // location
  mu = a[clade_id];
}
model {
  K ~ normal(mu, sigma);         // probability model
  sigma ~ uniform(0, 50);        // prior: scale
  for (i in 1:4) a[i] ~ normal(0, 0.5);
}
'
mod <- stan(model_code = model, data = dat)

summary(mod, 'a')$summary
```
