---
title: "Statistical Rethinking 2: Chapter 11"
author: Vincent Arel-Bundock
output: html_document
---

# Status

Estimated and checked against the book:

* m11.1
* m11.2
* m11.3
* m11.4
* m11.5
* m11.6
* m11.7
* m11.8
* m11.9
* m11.10
* m11.11
* m11.12
* m11.13
* m11.14
* m_binom
* m_pois

# Libraries

```{r, include=FALSE}
# set default graphics theme if Vincent's personal package is installed
if (requireNamespace('vincent', quietly = TRUE)) {
        vincent::theming()
}
```

```{r, message=FALSE}
library(tidyverse)
library(tidybayes)
library(rstan)
library(rethinking)
options(mc.cores = 4)
```

Was the `R` number generator different when the book was written? I know it changed recently.

# Section 11.1: Binomial regression

```{r, warning=FALSE, message=FALSE, results=FALSE}
chimpanzees <- read.csv('data/chimpanzees.csv', sep = ';') %>%
               mutate(treatment = factor(1 + prosoc_left + 2 * condition),
                      actor = factor(actor))

stan_data <- compose_data(chimpanzees,
                          n_actor = n_distinct(actor),
                          n_treatment = n_distinct(treatment),
                          n_predictions = n_actor * n_treatment)

stan_program <- '
data {
  int<lower=1> n;                       
  int<lower=0, upper=1> pulled_left[n];
}
parameters {
  real a;
}
transformed parameters{
  real<lower=0, upper=1> p;
  p = inv_logit(a);
}
model {
  a ~ normal(1, 10);
  pulled_left ~ binomial(1, p);
}
'

m11.1 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m11.1
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
stan_program <- '
data {
  int<lower=1> n;                       
  int<lower=1> n_treatment;
  int treatment[n];
  int<lower=0, upper=1> pulled_left[n];
}
parameters {
  real a;
  real b[n_treatment];
}
model {
  vector[n] p;
  for (i in 1:n) {
      p[i] = inv_logit(a + b[treatment[i]]);
  }
  pulled_left ~ binomial(1, p);
  a ~ normal(0, 1.5);
  b ~ normal(0, 10);
}
'

m11.2 <- stan(model_code = stan_program, data = stan_data, iter = 5000)
```

```{r}
m11.2
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
stan_program <- '
data {
  int<lower=1> n;                       
  int<lower=1> n_treatment;
  int treatment[n];
  int<lower=0, upper=1> pulled_left[n];
}
parameters {
  real a;
  real b[n_treatment];
}
model {
  vector[n] p;
  for (i in 1:n) {
      p[i] = inv_logit(a + b[treatment[i]]);
  }
  pulled_left ~ binomial(1, p);
  a ~ normal(0, 1.5);
  b ~ normal(0, 0.5);
}
'

m11.3 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m11.3
```

```{r, message=FALSE, fig.asp = .3}
stan_program <- '
data {
  int n;                       
  int n_treatment;                       
  int n_actor;
  int n_predictions;
  int actor[n];
  int pulled_left[n];
  int treatment[n];
}
parameters {
  real a[n_actor];
  real b[n_treatment];
}
transformed parameters{
  vector[n] p;
  for (i in 1:n) {
      p[i] = a[actor[i]] + b[treatment[i]];
      p[i] = inv_logit(p[i]);
  }
}
model {
  a ~ normal(0, 1.5);
  b ~ normal(0, 0.5);
  pulled_left ~ binomial(1, p);
}
generated quantities {
  real db13;
  real db24;
  db13 = b[1] - b[3];
  db24 = b[2] - b[4];
}
'

m11.4 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
summary(m11.4, c('db13', 'db24', 'a', 'b'))$summary
```

```{r}
datplot <- m11.4 %>% 
           spread_draws(a[actor]) %>%
           mean_qi() %>%
           mutate(actor = factor(actor, 7:1))
ggplot(datplot, aes(a, actor, xmin = .lower, xmax = .upper)) +
    geom_vline(xintercept = 0, linetype = 'dashed') +
    geom_pointrange()
```

```{r, fig.asp = .2}
datplot <- m11.4 %>% 
           spread_draws(b[treatment]) %>%
           mean_qi() %>%
           mutate(treatment = c("R/N", "L/N", "R/P", "L/P")[treatment])
ggplot(datplot, aes(b, treatment, xmin = .lower, xmax = .upper)) +
    geom_vline(xintercept = 0, linetype = 'dashed') +
    geom_pointrange()
```

```{r, fig.asp = .2}
datplot <- m11.4 %>% 
           gather_draws(db13, db24) %>%
           median_qi()
ggplot(datplot, aes(.value, .variable, xmin = .lower, xmax = .upper)) +
  geom_vline(linetype = 'dashed', xintercept = 0) +
  geom_pointrange() + 
  labs(x = 'Value', y = '')
```

```{r, fig.asp=.4, out.width='100%'}

tmp <- chimpanzees %>% mutate(i = 1:n())
datplot <- m11.4 %>% 
           spread_draws(p[i]) %>%
           mean_qi() %>%
           left_join(tmp, by = 'i')  %>%
           mutate(treatment = c('R/N', 'L/N', 'R/P', 'L/P')[treatment],
                  treatment = factor(treatment, levels = c('R/N', 'L/N', 'R/P', 'L/P')))

ggplot(datplot, aes(treatment, p, ymin = .lower, ymax = .upper, label = treatment)) +
  geom_text() +
  geom_hline(linetype = 'dashed', yintercept = 0.5) +
  geom_linerange(alpha = .1) + 
  facet_grid(~ actor) +
  labs(x = 'Value', y = '') +
  theme_classic()
```

```{r, warning=FALSE, message=FALSE, results=FALSE}

stan_data <- read.csv('data/chimpanzees.csv', sep = ';') %>%
             mutate(treatment = factor(1 + prosoc_left + 2 * condition),
                    side = prosoc_left + 1,
                    cond = condition + 1) %>%
             compose_data
stan_data$n_actor <- n_distinct(stan_data$actor)
stan_data$n_side <- n_distinct(stan_data$side)
stan_data$n_treatment <- n_distinct(stan_data$treatment)
stan_data$n_cond <- n_distinct(stan_data$cond)

stan_program <- '
data {
  int n;
  int n_actor;
  int n_treatment;
  int n_side;
  int treatment[n];
  int side[n];
  int actor[n];
  int pulled_left[n];
}
parameters {
  real a[n_actor];
  real bc[n_treatment];
  real bs[n_side];
}
model {
  vector[n] p;
  a ~ normal(0, 1.5);
  bs ~ normal(0, 0.5);
  bc ~ normal(0, 0.5);
  for (i in 1:n) {
      p[i] = inv_logit(a[actor[i]] + bc[treatment[i]] + bs[side[i]]);
  }
  pulled_left ~ binomial(1, p);
}
'
m11.5 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m11.5
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
chimpanzees <- read.csv('data/chimpanzees.csv', sep = ';') %>%
               mutate(treatment = factor(1 + prosoc_left + 2 * condition),
                      actor = factor(actor),
                      side = prosoc_left + 1,
                      cond = condition + 1) %>%
               group_by(treatment, actor, side, cond) %>%
               summarize(left_pulls = sum(pulled_left))
stan_data <- compose_data(chimpanzees)

stan_program <- '
data {
  int n;
  int n_actor;
  int n_treatment;
  int treatment[n];
  int actor[n];
  int left_pulls[n];
}
parameters {
  real a[n_actor];
  real b[n_treatment];
}
model {
  vector[n] p;
  a ~ normal(0, 1.5);
  b ~ normal(0, 0.5);
  for (i in 1:n) {
      p[i] = inv_logit(a[actor[i]] + b[treatment[i]]);
  }
  left_pulls ~ binomial(18, p);
}
'
m11.6 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m11.6
```

```{r, fig.asp = .7, out.width='100%'}
UCBadmit <- read.csv('data/UCBadmit.csv', sep = ';') %>%
            rename(rejection = reject,
                   applicant_gender = applicant.gender) %>%
            mutate(ratio = admit / applications)
stan_data <- compose_data(UCBadmit)

stan_program_A <- '
data {
  int n;
  int n_dept;
  int admit[n];
  int applications[n];
  int applicant_gender[n];
}
parameters {
  real a[2];
}
transformed parameters {
  vector[n] p;
  for (i in 1:n) {
    p[i] = inv_logit(a[applicant_gender[i]]);
  }
}
model {
  a ~ normal(0, 1.5);
  for (i in 1:n) {
    admit[i] ~ binomial(applications[i], p[i]);
  }
}
'

stan_program_B <- '
data {
  int n;
  int n_dept;
  int admit[n];
  int applications[n];
  int applicant_gender[n];
  int dept[n];
}
parameters {
  real a[2];
  real b[n_dept];
}
transformed parameters {
  vector[n] p;
  for (i in 1:n) {
    p[i] = inv_logit(a[applicant_gender[i]] + b[dept[i]]);
  }
}
model {
  a ~ normal(0, 1.5);
  b ~ normal(0, 1.5);
  for (i in 1:n) {
    admit[i] ~ binomial(applications[i], p[i]);
  }
}
'

m11.7 <- stan(model_code = stan_program_A, data = stan_data)
m11.8 <- stan(model_code = stan_program_B, data = stan_data)

tabA <- UCBadmit %>%
        mutate(`Model A` = summary(m11.7, 'p')$summary[, 1])
tabB <- UCBadmit %>%
        mutate(`Model B` = summary(m11.8, 'p')$summary[, 1])
tab <- left_join(tabA, tabB) %>%
       rename(Observed = ratio) %>% 
       pivot_longer(cols = c(6:8))

ggplot(tab, aes(applicant_gender, value, color = name, shape = applicant_gender), alpha = .8) +
    geom_point(size = 4) +
    facet_grid(. ~ dept) +
    labs(x = '', y = 'Probability of admission', color = '', shape = '')
```

```{r}
summary(m11.7, c('a'))$summary
summary(m11.8, c('a', 'b'))$summary
```

# Section 11.2: Poisson regression

```{r, fig.asp = .5}

kline <- read.csv('data/kline.csv', sep = ';') %>%
         mutate(P = as.vector(scale(log(population))))
stan_data <- compose_data(kline)

stan_program <- '
data {
  int n;
  int total_tools[n];
}
parameters {
  real a;
}
model {
  vector[n] lambda;
  a ~ normal(3, 0.5);
  for (i in 1:n) {
    lambda[i] = exp(a);
  }
  total_tools ~ poisson(lambda);
}
'
m11.9 <- stan(model_code = stan_program, data = stan_data)

stan_program <- '
data {
  int n;
  int n_contact;
  int contact[n];
  real P[n];
  real Pnew[100];
  int total_tools[n];
}
parameters {
  real a[n_contact];
  real b[n_contact];
}
model {
  vector[n] lambda;
  for (i in 1:n) {
    lambda[i] = exp(a[contact[i]] + b[contact[i]] * P[i]);
  }
  a ~ normal(3, 0.5);
  b ~ normal(0, 0.2);
  total_tools ~ poisson(lambda);
}
generated quantities {
  real yhat[100, 2];
  for (i in 1:100) {
    for (j in 1:2) {
      yhat[i, j] = exp(a[j] + b[j] * Pnew[i]);
    }
  }
}
'
stan_data$Pnew <- seq(min(stan_data$P), max(stan_data$P), length.out = 100)
m11.10 <- stan(model_code = stan_program, data = stan_data)

datplot <- m11.10 %>%
           gather_draws(yhat[idx, contact]) %>%
           median_qi() %>%
           left_join(tibble(idx = 1:100, 'Population' = stan_data$Pnew), by = 'idx') %>%
           mutate(contact = ifelse(contact == 1, 'Low', 'High')) %>%
           rename(Tools = .value)

ggplot(datplot, aes(Population, Tools, linetype = contact, ymax = .upper, ymin = .lower, fill = contact)) +
    geom_ribbon(alpha = .2) +
    geom_line() +
    xlab('log population (std)')
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
stan_program <- '
data {
  int n;
  int n_contact;
  int contact[n];
  real population[n];
  int total_tools[n];
}
parameters {
  real a[n_contact];
  real<lower=0> b[n_contact];
  real<lower=0> g;
}
model {
  vector[n] lambda;
  for (i in 1:n) {
    lambda[i] = exp(a[contact[i]]) * population[i]^b[contact[i]] / g;
  }
  a ~ normal(1, 1);
  b ~ exponential(1);
  g ~ exponential(1);
  total_tools ~ poisson(lambda);
}
'

m11.11 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m11.11
```

#### Section 11.2.2

```{r, fig.asp = .4, warning=FALSE, message=FALSE, results=FALSE}

old <- data.frame('monastery' = 0,
                  'exposure' = 1,
                  'y' = rpois(30, 1.5))
new <- data.frame('monastery' = 1,
                  'exposure' = 7,
                  'y' = rpois(4, 0.5 * 7))
monasteries <- rbind(old, new) %>%
               mutate(log_days = log(exposure))
stan_data <- compose_data(monasteries)

stan_program <- '
data {
  int<lower=1> n;
  real<lower=0> log_days[n];
  int<lower=0, upper=1> monastery[n];
  int y[n];
}
parameters {
  real a;
  real b;
}
model {
  real lambda[n];
  for (i in 1:n) {
    lambda[i] = exp(log_days[i] + a + b * monastery[i]);
  }
  y ~ poisson(lambda);
  a ~ normal(0, 1);
  b ~ normal(0, 1);
}
generated quantities {
  real Old;
  real New;
  Old = exp(a);
  New = exp(a + b);
}
'

m11.12 <- stan(model_code = stan_program, data = stan_data)

datplot <- m11.12 %>% 
           gather_draws(Old, New)

ggplot(datplot, aes(.value, .variable)) + 
    stat_halfeye() +
    labs(x = 'Posterior distribution of monastic productivity', y = '')
```

# Section 11.3: Multinomial and categorical models

Results for models m11.13 and m11.14 differ from the book but are identical to those obtained by running the `rethink` replication code. Perhaps something changed in `R` random number generator between the publication of the book and my replication.

```{r, warning=FALSE, message=FALSE, results=FALSE}
# simulate career choices among 500 individuals
N <- 500             # number of individuals
income <- c(1,2,5)   # expected income of each career
score <- 0.5*income  # scores for each career, based on income

# next line converts scores to probabilities
p <- softmax(score[1],score[2],score[3])

# now simulate choice
# outcome career holds event type values, not counts
career <- rep(NA,N)  # empty vector of choices for each individual

# sample chosen career for each individual
set.seed(34302)
for ( i in 1:N ) career[i] <- sample( 1:3 , size=1 , prob=p )
stan_data <- list( N=N , K=3 , career=career , career_income=income )

stan_program <- "
data{
    int N; // number of individuals
    int K; // number of possible careers
    int career[N]; // outcome
    vector[K] career_income;
}
parameters{
    vector[K-1] a; // intercepts
    real<lower=0> b; // association of income with choice
}
model{
    vector[K] p;
    vector[K] s;
    a ~ normal( 0 , 1 );
    b ~ normal( 0 , 0.5 );
    s[1] = a[1] + b*career_income[1];
    s[2] = a[2] + b*career_income[2];
    s[3] = 0; // pivot
    p = softmax( s );
    career ~ categorical( p );
}
"

m11.13 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m11.13
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
## R code 11.59
N <- 500
# simulate family incomes for each individual
family_income <- runif(N)
# assign a unique coefficient for each type of event
b <- c(-2,0,2)
career <- rep(NA,N)  # empty vector of choices for each individual
for ( i in 1:N ) {
    score <- 0.5*(1:3) + b*family_income[i]
    p <- softmax(score[1],score[2],score[3])
    career[i] <- sample( 1:3 , size=1 , prob=p )
}
stan_data <- list( N=N , K=3 , career=career , family_income=family_income )

stan_program <- "
data{
    int N; // number of observations
    int K; // number of outcome values
    int career[N]; // outcome
    real family_income[N];
}
parameters{
    vector[K-1] a; // intercepts
    vector[K-1] b; // coefficients on family income
}
model{
    vector[K] p;
    vector[K] s;
    a ~ normal(0,1.5);
    b ~ normal(0,1);
    for ( i in 1:N ) {
        for ( j in 1:(K-1) ) s[j] = a[j] + b[j]*family_income[i];
        s[K] = 0; // the pivot
        p = softmax( s );
        career[i] ~ categorical( p );
    }
}
"

m11.14 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m11.14
```

#### Section 11.3.3

```{r, warning=FALSE, message=FALSE, results=FALSE}
UCBadmit <- read.csv('data/UCBadmit.csv', sep = ';') %>%
            rename(rejection = reject) %>%
            janitor::clean_names() %>%
            mutate(ratio = admit / applications)
stan_data <- compose_data(UCBadmit)
stan_program <- '
data {
  int n;
  int admit[n];
  int applications[n];
}
parameters {
  real a;
}
transformed parameters {
  real p;
  p = inv_logit(a);
}
model {
  for (i in 1:n) {
    admit[i] ~ binomial(applications[i], p);
  }
}
'
m_binom <- stan(model_code = stan_program, data = stan_data)

stan_program <- '
data {
  int n;
  int admit[n];
  int rejection[n];
  int applications[n];
}
parameters {
  real lambda[2];
  real a[2];
}
model {
  for (i in 1:n) {
    admit[i] ~ poisson(exp(a[1]));
    rejection[i] ~ poisson(exp(a[2]));
  }
}
generated quantities {
  real p;
  p = exp(a[1]) / (exp(a[1]) + exp(a[2]));
}
'
m_pois <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m_binom
m_pois
```
