---
title: "Statistical Rethinking 2: Chapter 13"
author: Vincent Arel-Bundock
output: html_document
---

# Status

Estimated and checked against the book:

* m13.1
* m13.2
* m13.3
* m13.4
* m13.5
* m13.6
* m13.7
* m13.7nc
* m13.4nc

# Libraries

```{r, include=FALSE}
# set default graphics theme if Vincent's personal package is installed
if (requireNamespace('vincent', quietly = TRUE)) {
        vincent::theming()
}
```

```{r, results=FALSE, message=FALSE}
library(tidyverse)
library(tidybayes)
library(rstan)
library(rethinking)
library(patchwork)
options(mc.cores = 4)
```

# Section 13:1: Multilevel tadpoles

```{r, results=FALSE, message=FALSE, warning=FALSE}
dat <- read.csv('data/reedfrogs.csv', sep = ';')
stan_data <- list(n = nrow(dat),
                  surv = dat$surv,
                  dens = dat$density,
                  tank = 1:nrow(dat))

stan_program <- '
data {
    int n;
    int surv[n];
    int dens[n];
    int tank[n];
}
parameters {
    real a[n];
}
transformed parameters {
    vector[n] p;
    for (i in 1:n) {
        p[i] = inv_logit(a[i]);
    }
}
model {
    a ~ normal(0, 1.5);
    for (i in 1:n) {
        surv[i] ~ binomial(dens[i], p[i]);
    }
}
'

m13.1 <- stan(model_code = stan_program, data = stan_data)

stan_program <- '
data {
    int n;
    int surv[n];
    int dens[n];
    int tank[n];
}
parameters {
    real abar;
    real<lower=0> sigma;
    real a[n];
}
transformed parameters {
    vector[n] p;
    for (i in 1:n) {
        p[i] = inv_logit(a[i]);
    }
}
model {
    a ~ normal(abar, sigma);
    sigma ~ exponential(1);
    for (i in 1:n) {
        surv[i] ~ binomial(dens[i], p[i]);
    }
}
'

m13.2 <- stan(model_code = stan_program, data = stan_data)
```

```{r, out.width = '100%', fig.asp = .5}

datplot <- tibble(Tank = 1:nrow(dat),
                  Density = dat$density,
                  'Observed' = dat$propsurv,
                  `No pooling` = summary(m13.1, 'p')$summary[, 1],
                  `Partial pooling` = summary(m13.2, 'p')$summary[, 1]) %>%
           mutate(Size = case_when(Density == 10 ~ 'Small',
                                   Density == 25 ~ 'Medium',
                                   Density == 35 ~ 'Large'),
                  Size = factor(Size, c('Small', 'Medium', 'Large'))) %>%
           pivot_longer(-c(Tank, Size, Density))

ggplot(datplot, aes(Tank, value, color = name)) +
    geom_point(size = 2) +
    facet_grid(. ~ Size, scales = 'free_x') +
    ylab('Survival proportion')
```

```{r, out.width='100%', fig.asp = .6}

post <- m13.2 %>%
        spread_draws(abar, sigma)
datplot <- post %>%
		   sample_n(1000)

# side-by-side plots
par(mfrow = c(1, 2))

# log odds plot for 1000 samples
plot(NULL, xlim = c(-3, 4), ylim = c(0, .35),
     xlab = 'Log-odds survive', ylab = 'Density')
for (i in 1:nrow(datplot)) {
    curve(dnorm(x, datplot$abar[i], datplot$sigma[i]), add = TRUE,
          col = adjustcolor('black', alpha = .02))
}

# density plot
sim_tanks <- rnorm(1:nrow(post), post$abar, post$sigma)
sim_tanks <- density(rethinking::inv_logit(sim_tanks))
plot(sim_tanks, lwd = 2, adj = .1, xlab = 'Probability survive', main = '')
```

Results are different because of different random seeds. This model replicates the same results produced by the book's replication code:

```{r, warning=FALSE, message=FALSE, results=FALSE}
set.seed(5005)
a_bar <- 1.5
sigma <- 1.5
nponds <- 60
Ni <- as.integer(rep(c(5, 10, 25, 35), each = 15))
a_pond <- rnorm(nponds, mean = a_bar, sd = sigma)
dsim <- data.frame(pond = 1:nponds,
                   Ni = Ni,
                   true_a = a_pond)
dsim$Si <- rbinom(nponds, prob = logistic(dsim$true_a), size = dsim$Ni)
dsim$p_nopool <- dsim$Si / dsim$Ni
stan_data <- compose_data(dsim)
stan_data$n_pond <- n_distinct(stan_data$pond)

stan_program <- "
data {
    int n;
    int n_pond;
    int Si[n];
    int Ni[n];
    int pond[n];
}
parameters {
    real a_pond[n_pond];
    real a_bar;
    real<lower=0> sigma;
}
model {
    vector[n] p;
    for (i in 1:n) {
        p[i] = inv_logit(a_pond[pond[i]]);
    }
    Si ~ binomial(Ni, p);
    a_pond ~ normal(a_bar, sigma);
    a_bar ~ normal(0, 1.5);
    sigma ~ exponential(1);
}
"

m13.3 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
summary(m13.3, c('a_bar', 'sigma'))$summary
```


```

# Section 13.3: More than one type of cluster

```{r, results=FALSE, message=FALSE, warning=FALSE}
dat <- read.csv('data/chimpanzees.csv', sep = ';')
stan_data <- list(actor = dat$actor,
                  treatment = as.integer(1 + dat$prosoc_left + 2 * dat$condition),
                  pulled_left = dat$pulled_left,
                  block_id = dat$block,
                  n = nrow(dat),
                  n_block_id = length(unique(dat$block)))
stan_data$n_treatment <- length(unique(stan_data$treatment))

stan_program <- '
data {
    int n;
    int n_block_id;
    int n_treatment;
    int actor[n];
    int treatment[n];
    int pulled_left[n];
    int block_id[n];
}
parameters {
    vector[n] a;
    vector[n_block_id] g;
    vector[n_treatment] b;
    real a_bar;
    real<lower=0> sigma_a;
    real<lower=0> sigma_g;
}
transformed parameters {
    vector[n] p;
    for (i in 1:n) {
        p[i] = inv_logit(a[actor[i]] + g[block_id[i]] + b[treatment[i]]);
    }
}
model {
    a_bar ~ normal(0, 1.5);
    a ~ normal(a_bar, sigma_a);
    b ~ normal(0, 0.5);
    g ~ normal(0, sigma_g);
    sigma_a ~ exponential(1);
    sigma_g ~ exponential(1);
    pulled_left ~ binomial(1, p);
}
'

m13.4 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
summary(m13.4, c('a_bar', 'sigma_a', 'sigma_g'))$summary
```

```{r, results=FALSE, message=FALSE, warning=FALSE}
stan_program <- '
data {
    int n;
    int n_block_id;
    int n_treatment;
    int actor[n];
    int treatment[n];
    int pulled_left[n];
    int block_id[n];
}
parameters {
    vector[n] a;
    vector[n_block_id] g;
    vector[n_treatment] b;
    real a_bar;
    real<lower=0> sigma_a;
    real<lower=0> sigma_g;
}
transformed parameters {
    vector[n] p;
    p = inv_logit(a[actor] + b[treatment]);
}
model {
    a_bar ~ normal(0, 1.5);
    a ~ normal(a_bar, sigma_a);
    b ~ normal(0, 0.5);
    g ~ normal(0, sigma_g);
    sigma_a ~ exponential(1);
    sigma_g ~ exponential(1);
    pulled_left ~ binomial(1, p);
}
'

m13.5 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
summary(m13.5, c('a_bar', 'sigma_a', 'sigma_g', 'g', 'b'))$summary
```

```{r, results=FALSE, message=FALSE, warning = FALSE}
stan_program <- '
data {
    int n;
    int n_block_id;
    int n_treatment;
    int actor[n];
    int treatment[n];
    int pulled_left[n];
    int block_id[n];
}
parameters {
    vector[n] a;
    vector[n_block_id] g;
    vector[n_treatment] b;
    real a_bar;
    real<lower=0> sigma_a;
    real<lower=0> sigma_g;
    real<lower=0> sigma_b;
}
transformed parameters {
    vector[n] p;
    p = inv_logit(a[actor] + g[block_id] + b[treatment]);
}
model {
    a_bar ~ normal(0, 1.5);
    a ~ normal(a_bar, sigma_a);
    b ~ normal(0, sigma_b);
    g ~ normal(0, sigma_g);
    sigma_a ~ exponential(1);
    sigma_g ~ exponential(1);
    sigma_b ~ exponential(1);
    pulled_left ~ binomial(1, p);
}
'

m13.6 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
summary(m13.6, c('a_bar', 'sigma_a', 'sigma_g', 'sigma_b', 'g', 'b'))$summary
```

# Section 13: Divergent transitions and non-centered priors

```{r, results=FALSE, message=FALSE, warning=FALSE}
stan_program <- '
data {
  int n;
}
parameters {
  real v;
  real x;
}
model {
  x ~ normal(0, exp(v));
  v ~ normal(0, 3);
}
'
stan_data <- list(n = 1)
m13.7 <- stan(model_code = stan_program, data = stan_data)
```
```{r}
m13.7
```

```{r, results=FALSE, message=FALSE, warning=FALSE}
stan_program <- '
data {
  int n;
}
parameters {
  real v;
  real z;
}
model {
  z ~ normal(0, 1);
  v ~ normal(0, 3);
}
generated quantities {
  real x;
  x = z * exp(v);
}
'

m13.7nc <- stan(model_code = stan_program, data = stan_data, iter = 5000)
```

```{r}
m13.7nc
```

```{r, results=FALSE, message=FALSE}
stan_program <- '
data {
    int n;
    int n_block_id;
    int n_treatment;
    int n_actor;
    int actor[n];
    int treatment[n];
    int pulled_left[n];
    int block_id[n];
}
parameters {
  vector[n_treatment] b;
  vector[n_actor] z;
  vector[n_block_id] x;
  real<lower=0> sigma_a;
  real<lower=0> sigma_g;
  real abar;
}
transformed parameters {
  vector[n] a;
  vector[n] g;
  vector[n] p;
  for (i in 1:n) {
      a[i] = abar + z[actor[i]] * sigma_a;
      g[i] = x[block_id[i]] * sigma_g;
      p[i] = a[i] + g[i] + b[treatment[i]];
      p[i] = inv_logit(p[i]);
  }
}
model {
  pulled_left ~ binomial(1, p);
  b ~ normal(0, 0.5);
  z ~ normal(0, 1);
  x ~ normal(0, 1);
  abar ~ normal(0, 1.5);
  sigma_a ~ exponential(1);
  sigma_g ~ exponential(1);
}
'

dat <- read.csv('data/chimpanzees.csv', sep = ';')
stan_data <- list(actor = dat$actor,
                  treatment = as.integer(1 + dat$prosoc_left + 2 * dat$condition),
                  pulled_left = dat$pulled_left,
                  block_id = dat$block,
                  n = nrow(dat),
                  n_block_id = length(unique(dat$block)),
                  n_actor = length(unique(dat$actor)))
stan_data$n_treatment <- length(unique(stan_data$treatment))


m13.4nc <- stan(model_code = stan_program, data = stan_data)
```

```{r}
summary(m13.4nc, c('b', 'abar', 'sigma_a', 'sigma_g'))$summary
```


