---
title: "Statistical Rethinking 2: Chapter 14"
author: Vincent Arel-Bundock
output: html_document
---

# Status

Estimated and checked against the book:

* m14.1
* m14.2
* m14.3
* m14.4
* m14.5
* m14.6
* m14.7
* m14.8
* m14.9
* m14.10
* m14.11

# Libraries

```{r, include=FALSE}
# set default graphics theme if Vincent's personal package is installed
if (requireNamespace('vincent', quietly = TRUE)) {
        vincent::theming()
}
```

```{r, results=FALSE, message=FALSE, warning=FALSE}
library(tidybayes)
library(rstan)
library(patchwork)
library(rethinking)
library(tidyverse)
library(ape)
options(mc.cores = 4)
```

# Section 14.1: Varying slopes by construction

```{r, warning=FALSE, message=FALSE, results=FALSE}
# Simulation code from McElreath's replication files
library(MASS)
a <- 3.5            # average morning wait time
b <- (-1)           # average difference afternoon wait time
sigma_a <- 1        # std dev in intercepts
sigma_b <- 0.5      # std dev in slopes
rho <- (-0.7)       # correlation between intercepts and slopes
Mu <- c( a , b )
cov_ab <- sigma_a*sigma_b*rho
Sigma <- matrix( c(sigma_a^2,cov_ab,cov_ab,sigma_b^2) , ncol=2 )
sigmas <- c(sigma_a,sigma_b) # standard deviations
Rho <- matrix( c(1,rho,rho,1) , nrow=2 ) # correlation matrix
Sigma <- diag(sigmas) %*% Rho %*% diag(sigmas)
N_cafes <- 20
set.seed(5) # used to replicate example
vary_effects <- mvrnorm( N_cafes , Mu , Sigma )
a_cafe <- vary_effects[,1]
b_cafe <- vary_effects[,2]
set.seed(22)
N_visits <- 10
afternoon <- rep(0:1,N_visits*N_cafes/2)
cafe_id <- rep( 1:N_cafes , each=N_visits )
mu <- a_cafe[cafe_id] + b_cafe[cafe_id]*afternoon
sigma <- 0.5  # std dev within cafes
wait <- rnorm( N_visits*N_cafes , mu , sigma )
d <- data.frame( cafe=cafe_id , afternoon=afternoon , wait=wait )

stan_data <- compose_data(d,
 		                  n_cafe = n_distinct(cafe))

stan_program <- "
data {
  int n;
  int n_cafe;
  int cafe[n];
  vector[n] afternoon;
  vector[n] wait;
}
parameters {
  vector[n_cafe] a_cafe;
  vector[n_cafe] b_cafe;
  real a;
  real b;
  vector<lower=0>[2] sigma_cafe;
  real<lower=0> sigma;
  corr_matrix[2] Rho;
}
model {
  vector[n] mu;
  vector[2] YY[n_cafe];
  vector[2] MU;
  Rho ~ lkj_corr(2);
  sigma ~ exponential(1);
  sigma_cafe ~ exponential(1);
  a ~ normal(5, 2);
  b ~ normal(-1, .5);
  MU = [a, b]';
  for (j in 1:n_cafe) {
    YY[j] = [a_cafe[j], b_cafe[j]]';
  }
  YY ~ multi_normal(MU, quad_form_diag(Rho, sigma_cafe));
  mu = a_cafe[cafe] + b_cafe[cafe] .* afternoon;
  wait ~ normal(mu, sigma);
}
"

m14.1 <- stan(model_code = stan_program, data = stan_data)

datplot <- m14.1 %>% 
           spread_draws(Rho[i, j]) %>%
           filter(i == 1, j == 2)

ggplot(datplot, aes(Rho)) +
    geom_density() +
    xlim(-1, 1) +
    xlab('Correlation')
```

# Section 14.2: Advanced varying slopes

```{r, warning=FALSE, message=FALSE, results=FALSE}

stan_data <- read.csv('data/chimpanzees.csv', sep = ';') %>%
             mutate(treatment = 1 + prosoc_left + 2 * condition) %>%
             rename(block_id = block) %>%
             compose_data(n_treatment = n_distinct(treatment),
                          n_actor = n_distinct(actor),
                          n_block_id = n_distinct(block_id))

stan_program <- "
data {
    int n;
    int n_treatment;
    int n_actor;
    int n_block_id;
    int pulled_left[n];
    int actor[n];
    int treatment[n];
    int block_id[n];
}
parameters {
   vector[n_treatment] g; 
   vector[n_treatment] alpha[n_actor];
   vector[n_treatment] beta[n_block_id];
   vector<lower=0>[4] sigma_actor;
   vector<lower=0>[4] sigma_block;
   corr_matrix[4] Rho_block;
   corr_matrix[4] Rho_actor;
}
model {
    // probability model
    vector[n] p;
    for (i in 1:n) {
        p[i] = g[treatment[i]] + 
               alpha[actor[i], treatment[i]] + 
               beta[block_id[i], treatment[i]];
        p[i] = inv_logit(p[i]);
    }
    pulled_left ~ binomial(1, p);

    // adaptive priors
    beta ~ multi_normal(rep_vector(0, 4),
                        quad_form_diag(Rho_block, sigma_block));
    alpha ~ multi_normal(rep_vector(0, 4),
                         quad_form_diag(Rho_actor, sigma_actor));

    // fixed priors
    g ~ normal(0, 1);
    Rho_actor ~ lkj_corr(4);
    Rho_block ~ lkj_corr(4);
    sigma_actor ~ exponential(1);
    sigma_block ~ exponential(1);
}
"

m14.2 <- stan(model_code = stan_program, data = stan_data)
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
stan_program <- "
data {
    int n;
    int n_treatment;
    int n_actor;
    int n_block_id;
    int pulled_left[n];
    int actor[n];
    int treatment[n];
    int block_id[n];
}
parameters {
   matrix[n_treatment, n_actor] z_actor;
   matrix[n_treatment, n_block_id] z_block;
   vector[n_treatment] g; 
   vector<lower=0>[n_treatment] sigma_actor;
   vector<lower=0>[n_treatment] sigma_block;
   cholesky_factor_corr[n_treatment] L_Rho_block;
   cholesky_factor_corr[n_treatment] L_Rho_actor;
}
transformed parameters {
   matrix[n_actor, n_treatment] alpha;
   matrix[n_block_id, n_treatment] beta;
   beta = (diag_pre_multiply(sigma_block, L_Rho_block) * z_block)';
   alpha = (diag_pre_multiply(sigma_actor, L_Rho_actor) * z_actor)';
}
model{
    vector[n] p;
    L_Rho_block ~ lkj_corr_cholesky( 2 );
    sigma_block ~ exponential( 1 );
    L_Rho_actor ~ lkj_corr_cholesky( 2 );
    sigma_actor ~ exponential( 1 );
    g ~ normal( 0 , 1 );
    to_vector( z_block ) ~ normal( 0 , 1 );
    to_vector( z_actor ) ~ normal( 0 , 1 );
    for ( i in 1:n ) {
        p[i] = g[treatment[i]] + alpha[actor[i], treatment[i]] + beta[block_id[i], treatment[i]];
        p[i] = inv_logit(p[i]);
    }
    pulled_left ~ binomial( 1 , p );
}
"

m14.3 <- stan(model_code = stan_program, data = stan_data)
```

# Section 14:3: Instruments and causal designs

Zero true effect of E is confounded:

```{r, results=FALSE, message=FALSE, warning=FALSE}
set.seed(73)
N <- 500
U_sim <- rnorm( N )
Q_sim <- sample( 1:4 , size=N , replace=TRUE )
E_sim <- rnorm( N , U_sim + Q_sim )
W_sim <- rnorm( N , U_sim + 0*E_sim )
stan_data <- list(
    W = as.vector(scale(W_sim)),
    E = as.vector(scale(E_sim)),
    Q = as.vector(scale(Q_sim)),
    N = N)

stan_program <- '
data{
    int N;
    vector[N] W;
    vector[N] E;
    vector[N] Q;
}
parameters {
    real aW;
    real bEW;
    real<lower=0> sigma;
}
model {
    vector[N] mu;
    mu = aW + bEW * E;
    W ~ normal(mu, sigma);
    aW ~ normal(0, .2);
    bEW ~ normal(0, .5);
    sigma ~ exponential(1);
}
'

m14.4 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m14.4
```

Controlling for the instrument leads to disaster:

```{r, results=FALSE, message=FALSE, warning=FALSE}
stan_program <- '
data{
    int N;
    vector[N] W;
    vector[N] E;
    vector[N] Q;
}
parameters {
    real aW;
    real bEW;
    real bQW;
    real<lower=0> sigma;
}
model {
    vector[N] mu;
    mu = aW + bEW * E + bQW * Q;
    W ~ normal(mu, sigma);
    aW ~ normal(0, .2);
    bEW ~ normal(0, .5);
    bQW ~ normal(0, .5);
    sigma ~ exponential(1);
}
'

m14.5 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m14.5
```

Instrumental variable model:

```{r, results=FALSE, message=FALSE, warning=FALSE}
stan_program <- "
data{
    int N;
    vector[N] W;
    vector[N] E;
    vector[N] Q;
}
parameters {
    real aE;
    real aW;
    real bEW;
    real bQE;
    corr_matrix[2] Rho;
    vector<lower=0>[2] Sigma;
}
model {
    vector[2] MU[N];
    vector[2] YY[N];
    vector[N] mu_w;
    vector[N] mu_e;
    Sigma ~ exponential(1);
    Rho ~ lkj_corr(2);
    bQE ~ normal(0, .5);
    bEW ~ normal(0, .5);
    aW ~ normal(0, .2);
    aE ~ normal(0, .2);
    for (j in 1:N) {
        mu_w[j] = aW + bEW * E[j];
        mu_e[j] = aE + bQE * Q[j];
    }
    for (j in 1:N) {
        MU[j] = [mu_w[j], mu_e[j]]';
        YY[j] = [W[j], E[j]]';
    }
    YY ~ multi_normal(MU, quad_form_diag(Rho, Sigma));
}
"

m14.6 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m14.6
```

### m14.7

```{r}

load('data/KosterLeckie.rda') 
stan_data <- kl_dyads %>%
             compose_data(n_households = max(hidB))

stan_program <- "
data {
  int n;
  int n_households;
  int hidA[n];
  int hidB[n];
  int giftsAB[n];
  int giftsBA[n];
  int did[n];
}
parameters{
    real a;
    vector[2] gr[n_households];
    corr_matrix[2] Rho_gr;
    vector<lower=0>[2] sigma_gr;
    matrix[2,n] z;
    cholesky_factor_corr[2] L_Rho_d;
    real<lower=0> sigma_d;
}
transformed parameters{
    matrix[n,2] d;
    d = (diag_pre_multiply(rep_vector(sigma_d, 2), L_Rho_d) * z)';
}
model{
    vector[n] lambdaAB;
    vector[n] lambdaBA;
    sigma_d ~ exponential( 1 );
    L_Rho_d ~ lkj_corr_cholesky( 8 );
    to_vector( z ) ~ normal( 0 , 1 );
    sigma_gr ~ exponential( 1 );
    Rho_gr ~ lkj_corr( 4 );
    gr ~ multi_normal( rep_vector(0,2) , quad_form_diag(Rho_gr , sigma_gr) );
    a ~ normal( 0 , 1 );
    for ( i in 1:n ) {
        lambdaBA[i] = a + gr[hidB[i], 1] + gr[hidA[i], 2] + d[did[i], 2];
        lambdaBA[i] = exp(lambdaBA[i]);
        lambdaAB[i] = a + gr[hidA[i], 1] + gr[hidB[i], 2] + d[did[i], 1];
        lambdaAB[i] = exp(lambdaAB[i]);
    }
    giftsBA ~ poisson( lambdaBA );
    giftsAB ~ poisson( lambdaAB );
}
"
m14.7 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
summary(m14.7, c('Rho_gr', 'sigma_gr'))$summary
```

# Section 14.5: Continuous categories and the Gaussian process

```{r, warning=FALSE, message=FALSE, results=FALSE}
load('data/islandsDistMatrix.rda')
stan_data <- read.csv('data/Kline2.csv', sep = ';') %>%
             mutate(society = 1:10) %>%
             compose_data(Dmat = islandsDistMatrix)

stan_program <- "
// cov_GPL2 macro extracted from ulam object with get_stancode
functions{
    matrix cov_GPL2(matrix x, real sq_alpha, real sq_rho, real delta) {
        int N = dims(x)[1];
        matrix[N, N] K;
        for (i in 1:(N-1)) {
          K[i, i] = sq_alpha + delta;
          for (j in (i + 1):N) {
            K[i, j] = sq_alpha * exp(-sq_rho * square(x[i,j]) );
            K[j, i] = K[i, j];
          }
        }
        K[N, N] = sq_alpha + delta;
        return K;
    }
}
data {
    int n;
    int total_tools[n];
    int population[n];
    int society[n];
    matrix[n, n] Dmat;
}
parameters {
    vector[n] k;
    real<lower=0> a;
    real<lower=0> b;
    real<lower=0> g;
    real<lower=0> etasq;
    real<lower=0> rhosq;
}
model{
    vector[n] lambda;
    matrix[n, n] SIGMA;
    rhosq ~ exponential( 0.5 );
    etasq ~ exponential( 2 );
    a ~ exponential( 1 );
    b ~ exponential( 1 );
    g ~ exponential( 1 );
    SIGMA = cov_GPL2(Dmat, etasq, rhosq, 0.01);
    k ~ multi_normal( rep_vector(0,n) , SIGMA );
    for ( i in 1:n ) {
        lambda[i] = (a * population[i]^b/g) * exp(k[society[i]]);
    }
    total_tools ~ poisson( lambda );
}
"

m14.8 <- stan(model_code = stan_program, data = stan_data)
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
standardize <- function(x) as.vector(scale(x))
dat <- read.csv('data/Primates301.csv', sep = ';') %>%
             dplyr::select(name, mass = body, brain, group_size) %>%
             drop_na %>%
             mutate(mass = standardize(log(mass)),
                    brain = standardize(log(brain)),
                    group_size = standardize(log(group_size))) 
stan_data <- dat %>%
             compose_data(Dmat = diag(nrow(.)))

stan_program <- "
data{
    int n;
    vector[n] mass;
    vector[n] brain;
    vector[n] group_size;
    matrix[n, n] Dmat;
}
parameters{
    real a;
    real bG;
    real bM;
    real<lower=0> sigma_sq;
}
model{
    vector[n] mu;
    matrix[n, n] S;
    sigma_sq ~ exponential( 1 );
    bM ~ normal( 0 , 0.5 );
    bG ~ normal( 0 , 0.5 );
    a ~ normal( 0 , 1 );
    S = Dmat * sigma_sq;
    for ( i in 1:n ) {
        mu[i] = a + bM * mass[i] + bG * group_size[i];
    }
    brain ~ multi_normal( mu , S);
}
"

m14.9 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m14.9
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
library(ape)
data(Primates301_nex)

tree_trimmed <- keep.tip( Primates301_nex, dat$name )
Rbm <- corBrownian( phy=tree_trimmed )
V <- vcv(Rbm)
Dmat <- cophenetic( tree_trimmed )

stan_data <- compose_data(dat,
                          V = V[dat$name, dat$name],
                          Dmat = V / max(V))

m14.10 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m14.10
```

```{r, warning=FALSE, message=FALSE, results=FALSE}

stan_program <- "
functions{
    matrix cov_GPL1(matrix x, real sq_alpha, real sq_rho, real delta) {
        int N = dims(x)[1];
        matrix[N, N] K;
        for (i in 1:(N-1)) {
          K[i, i] = sq_alpha + delta;
          for (j in (i + 1):N) {
            K[i, j] = sq_alpha * exp(-sq_rho * x[i,j] );
            K[j, i] = K[i, j];
          }
        }
        K[N, N] = sq_alpha + delta;
        return K;
    }
}
data{
    int n;
    vector[n] mass;
    vector[n] brain;
    vector[n] group_size;
    matrix[n,n] Dmat;
}
parameters{
    real a;
    real bG;
    real bM;
    real<lower=0> etasq;
    real<lower=0> rhosq;
}
model{
    vector[n] mu;
    matrix[n, n] SIGMA;
    rhosq ~ normal( 3 , 0.25 );
    etasq ~ normal( 1 , 0.25 );
    bM ~ normal( 0 , 0.5 );
    bG ~ normal( 0 , 0.5 );
    a ~ normal( 0 , 1 );
    SIGMA = cov_GPL1(Dmat, etasq, rhosq, 0.01);
    for ( i in 1:n ) {
        mu[i] = a + bM * mass[i] + bG * group_size[i];
    }
    brain ~ multi_normal( mu , SIGMA );
}
"

stan_data$Dmat <- Dmat[dat$name, dat$name] / max(Dmat)
m14.11 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m14.11
```
