---
title: "Statistical Rethinking 2: Chapter 15"
author: Vincent Arel-Bundock
output: html_document
---

# Status

Estimated and checked against book:

* m15.1
* m15.2
* m15.3
* m15.4
* m15.5

TODO:

* m15.6
* m15.7
* m15.8

# Libraries

```{r, include=FALSE}
# set default graphics theme if Vincent's personal package is installed
if (requireNamespace('vincent', quietly = TRUE)) {
        vincent::theming()
}
```

```{r, results=FALSE, message=FALSE}
library(tidyverse)
library(tidybayes)
library(rstan)
library(patchwork)
library(rethinking)
options(mc.cores = 4)
```

# Section 15.1: Measurement error

```{r, warning=FALSE, message=FALSE, results=FALSE}
d <- read.csv('data/WaffleDivorce.csv', sep = ';')
stan_data <- list(
    D_obs = standardize( d$Divorce ),
    D_sd = d$Divorce.SE / sd( d$Divorce ),
    M = standardize( d$Marriage ),
    A = standardize( d$MedianAgeMarriage ),
    N = nrow(d)
)

stan_program <- "
data {
    int N;
    vector[N] D_obs;
    vector[N] D_sd;
    vector[N] M;
    vector[N] A;
}
parameters {
    vector[N] D_true;
    real a;
    real bA;
    real bM;
    real<lower=0> sigma;
}
model {
    vector[N] mu;
    mu = a + bA * A + bM * M;
    D_true ~ normal(mu, sigma);
    D_obs ~ normal(D_true, D_sd);
    a ~ normal(0, .2);
    bA ~ normal(0, .5);
    bM ~ normal(0, .5);
    sigma ~ exponential(1);
}
"

m15.1 <- stan(model_code = stan_program, data = stan_data)
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
stan_data <- list(
    D_obs = standardize( d$Divorce ),
    D_sd = d$Divorce.SE / sd( d$Divorce ),
    M_obs = standardize( d$Marriage ),
    M_sd = d$Marriage.SE / sd( d$Marriage ),
    A = standardize( d$MedianAgeMarriage ),
    N = nrow(d)
)

stan_program <- "
data {
    int N;
    vector[N] D_obs;
    vector[N] D_sd;
    vector[N] M_obs;
    vector[N] M_sd;
    vector[N] A;
}
parameters {
    vector[N] D_true;
    real a;
    real bA;
    real bM;
    real<lower=0> sigma;
    real M_true[N];
}
model {
    vector[N] mu;
    for (i in 1:N) {
        mu[i] = a + bA * A[i] + bM * M_true[i];
    }
    D_true ~ normal(mu, sigma);
    D_obs ~ normal(D_true, D_sd);
    M_true ~ normal(0, 1);
    M_obs ~ normal(M_true, M_sd);
    a ~ normal(0, .2);
    bA ~ normal(0, .5);
    bM ~ normal(0, .5);
    sigma ~ exponential(1);
}
"

m15.2 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
datpost <- m15.2 %>%
           spread_draws(M_true[i], D_true[i]) %>%
           mean_qi %>%
           select(Marriage = M_true, Divorce = D_true)
datreal <- tibble(Marriage = stan_data$M_obs, Divorce = stan_data$D_obs)
ggplot(datpost) +
    geom_point(aes(Marriage, Divorce), color = 'blue') +
    geom_point(data = datreal, aes(Marriage, Divorce), color = 'black')
```

# Section 15.2: Missing data

```{r, warning=FALSE, message=FALSE, results=FALSE}

## R code 15.7
N <- 500
A <- rnorm(N)
M <- rnorm(N,-A)
D <- rnorm(N,A)
A_obs <- rnorm(N,A)

## R code 15.8
N <- 100
S <- rnorm( N )
H <- rbinom( N , size=10 , inv_logit(S) )

## R code 15.9
D <- rbern( N ) # dogs completely random
Hm <- H
Hm[D==1] <- NA

## R code 15.10
D <- ifelse( S > 0 , 1 , 0 )
Hm <- H
Hm[D==1] <- NA

set.seed(501)
N <- 1000
X <- rnorm(N)
S <- rnorm(N)
H <- rbinom( N , size=10 , inv_logit( 2 + S - 2*X ) )
D <- ifelse( X > 1 , 1 , 0 )
Hm <- H
Hm[D==1] <- NA

## R code 15.12
stan_data <- list(
    H = H,
    S = S,
    n = length(H))

stan_program <- "
data{
    int n;
    int H[n];
    vector[n] S;
}
parameters {
    real a;
    real bS;
}
model {
    vector[n] p;
    p = a + bS * S;
    p = inv_logit(p);
    H ~ binomial(10, p);
    bS ~ normal(0, .5);
    a ~ normal(0, 1);
}
"

m15.3 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m15.3
```

```{r, warning=FALSE, message=FALSE, results=FALSE}
stan_data <- list(H = H[D == 0], S = S[D == 0], n = length(H[D == 0]))

stan_program <- "
data{
    int n;
    int H[n];
    vector[n] S;
}
parameters {
    real a;
    real bS;
}
model {
    vector[n] p;
    p = a + bS * S;
    p = inv_logit(p);
    H ~ binomial(10, p);
    bS ~ normal(0, .5);
    a ~ normal(0, 1);
}
"

m15.4 <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m15.4
```

Stan cannot handle `NA`s. Instead, we replace `NA`s by `Inf`, and we check for positive infinity in a loop to construct a `B_merge` vector. I'm sure there is a more computationally efficient (but convoluted?) way to achieve the same goal, but this hack is transparent and easy to understand.

```{r, warning=FALSE, message=FALSE, results=FALSE}
stan_data <- read.csv('data/milk.csv', sep = ';') %>%
             mutate(neocortex.prop = neocortex.perc / 100,
                    logmass = log(mass),
                    K = standardize(kcal.per.g),
                    M = standardize(logmass),
                    B = standardize(neocortex.prop),
                    B = ifelse(is.na(B), Inf, B)) %>%
             compose_data

stan_program <- "
data {
    int n;
    vector[n] K;
    vector[n] M;
    vector[n] B;
}
parameters {
    real nu;
    real a;
    real bB;
    real bM;
    real<lower=0> sigma_B;
    real<lower=0> sigma;
    vector[n] B_miss;
}
transformed parameters {
    vector[n] B_merge;
    for (i in 1:n) {
        if (B[i] == positive_infinity()) {
            B_merge[i] = B_miss[i];
        } else {
            B_merge[i] = B[i];
        }
    }
}
model {
    vector[n] mu;
    [sigma, sigma_B] ~ exponential(1);
    [a, nu, bB, bM] ~ normal(0, 0.5);
    B_merge ~ normal(nu, sigma_B);
    for (i in 1:n) {
        mu[i] = a + bB * B_merge[i] + bM * M[i];
    }
    K ~ normal(mu, sigma);
}
"
m15.5 <- stan(model_code = stan_program, data = stan_data,
            iter = 5000)

```{r}
summary(m15.5, c('nu', 'a', 'bM', 'bB', 'sigma_B', 'sigma'))$summary
```
